---
title: "P53 Data Analysis"
author: "Alyssa Rogers-Armstrong"
date: "February 2024"
output: pdf_document
---

```{r}
Data = read.csv("p53.csv")
View(Data)
library(ggplot2)
library(magrittr)
library(dplyr)
library(readr)
library(dplyr)
library(ggplot2)
library(pscl)
library(pROC)
library(plotROC)
library(caret)
```


## Exploratory Analysis


```{r}
# Display the structure of the data
str(Data)

# Summary statistics
summary(Data)

# Histogram of the percentage of nuclei stained with p53 in IHC
ggplot(Data, aes(x = Percent.Nuclei.Stained.in.p53.IHC)) +
  geom_histogram(binwidth = 0.1, fill = "skyblue", color = "black") +
  labs(title = "Distribution of % Nuclei Stained with p53 in IHC",
       x = "% Nuclei Stained", y = "Frequency")

# Boxplot of % Nuclei Stained by TP53 Sanger Status
ggplot(Data, aes(x = as.factor(TP53.Sanger.Status), y = Percent.Nuclei.Stained.in.p53.IHC)) +
  geom_boxplot(fill = "lightgreen", color = "black") +
  labs(title = "Boxplot of % Nuclei Stained by TP53 Sanger Status",
       x = "TP53 Sanger Status", y = "% Nuclei Stained")


# Outlier detection
outliers <- boxplot.stats(Data$Percent.Nuclei.Stained.in.p53.IHC)$out
print(outliers)

# Missing values
missing_values <- colSums(is.na(Data))
print(missing_values)


# Scatterplot
ggplot(Data, aes(x=proportion.of.nuclei.stained.with.p53.in.IHC, y=TP53.Sanger.Status..0.1.))+geom_point()+theme_bw()+
  labs(x="Proportion of Nuclei Stains (IHC)", y="Sanger Status")

# Scatterplot with Log Curve

# Predict probabilities using the logistic regression model
predicted_probs <- predict(logmodel, type = "response")

# Create scatterplot of data
plot(Data$Percent.Nuclei.Stained.in.p53.IHC, Data$TP53.Sanger.Status..0.1., 
     xlab = "Percent Nuclei Stained in p53 IHC", 
     ylab = "TP53 Mutation Status", 
     main = "Scatterplot with Logistic Regression Line")

# Add logistic regression line to the plot
x_values <- seq(min(Data$Percent.Nuclei.Stained.in.p53.IHC), max(Data$Percent.Nuclei.Stained.in.p53.IHC), length.out = 100)
y_values <- predict(logmodel, newdata = data.frame(Percent.Nuclei.Stained.in.p53.IHC = x_values), type = "response")
lines(x_values, y_values, col = "red")

ggplot(Data, aes(x=Percent.Nuclei.Stained.in.p53.IHC, y=TP53.Sanger.Status..0.1.,shape=TP53.Sanger.Status, color=TP53.Sanger.Status))+
  geom_point()+geom_jitter(height=0.09,width=0.09)+
  xlab("Percent Nuclei Stained in p53 IHC Test")+
  ylab("IHC Test Results")+
  ggtitle("Percent Nuclei Stained vs. IHC Test Results at 40% Threshold by TP53 Mutation")+
  stat_smooth(method='glm', color='blue', se=FALSE, method.args=list(family="binomial"))


library(ggplot2)
ggplot(Data, aes(x = Percent.Nuclei.Stained.in.p53.IHC, y = TP53.Sanger.Status..0.1., color=TP53.Sanger.Status..0.1.)) +
  geom_point() + # Add scatter plot points
  geom_smooth(method = "glm", method.args = list(family = "binomial"), se = FALSE) +  # Overlay logistic regression line
  labs(x = "Percent Nuclei Stained", y = "TP53 Sanger Status (0/1)") +  # Set axis labels
  theme_minimal() +  # Use minimal theme for cleaner appearance
  theme(legend.position="none")



ggplot(Data, aes(x=Percent.Nuclei.Stained.in.p53.IHC, y=TP53.Sanger.Status..0.1., shape=TP53.Sanger.Status, color=TP53.Sanger.Status))+
  geom_point(position=position_dodge(width=1.5))+
  xlab("Percent Nuclei Stained in p53 IHC Test")+
  ylab("Probability of TP53 Mutation")+
  ggtitle("Percent Nuclei Stained vs. Probability of TP53 Mutation")+
  theme(plot.title=element_text(hjust=0.1))+
  labs(shape="Actual TP53 Mutation", color="Actual TP53 Mutation")+
  theme(legend.position="bottom")+
  stat_smooth(method='glm', color='darkgreen',linewidth=0.5,
              se=FALSE,method.args=list(family=binomial))
```

## Statistical Analysis

Creating Factors:

```{r}
Data$TP53.Sanger.Status = as.factor(Data$TP53.Sanger.Status)
Data$TP53.Sanger.Status..0.1. = as.factor(Data$TP53.Sanger.Status..0.1.)
Data$IHC.P53.Status = as.factor(Data$IHC.P53.Status)
```


## Fit the Model:

```{r}
# Fit the logistic regression model
logmodel <- glm(TP53.Sanger.Status..0.1. ~ Percent.Nuclei.Stained.in.p53.IHC, data=Data, family = binomial)
summary(logmodel)
```

- A one unit increase in Percent of Nuclei Stained is associated with an average increase of 0.12120 in the log odds of TP53 Sanger Status.

## Calculating McFadden's R-Square
```{r}
# Calculate McFadden's R-Squared

pscl::pR2(logmodel)["McFadden"]
```

- We can compute a metric known as McFadden’s R2, which ranges from 0 to just under 1. Values close to 0 indicate that the model has no predictive power. In practice, values over 0.40 indicate that a model fits the data very well.

- McFadden = 0.7637797

- A value of 0.7637797 is very high for McFadden’s R2, which indicates that our model fits the data very well and has high predictive power.


## Odds Ratio

- Model Interpretation: Percent Nuclei Stained (p = 0.00633) is significantly associated with the TP53 Sanger Status at the 0.05 level. 

- Odds Ratio Interpretation: For each additional percentage of P53 nuclei stained in IHC, there is a 12.8% increased odds of a TP53 mutation.
```{r}
or_logmodel = exp(logmodel$coefficients)
ci_logmodel = exp(confint(logmodel))
orci_logmodel = cbind(or_logmodel, ci_logmodel)
orci_logmodel
```
## Confusion Matrix at 10% Threshold

- This is telling us that at the 10% threshold, there are a lot of false positives. This would waste a lot of resources by further testing these patients when they are true negatives.

```{r}
# Define the threshold
threshold <- 10

# Classify observations based on the threshold
predicted_TP53_status <- ifelse(Data$Percent.Nuclei.Stained.in.p53.IHC >= threshold, 1, 0)

# Create confusion matrix
conf_matrix <- table(Actual = Data$TP53.Sanger.Status..0.1., Predicted = predicted_TP53_status)


conf_matrix_obj <- as.matrix(conf_matrix)

# Create confusion matrix using caret
cm <- confusionMatrix(conf_matrix_obj)

# Convert to data frame for ggplot2
cm_df <- as.data.frame(cm$table)

cm_df$Actual <- ifelse(cm_df$Actual == 0, "Negative", "Positive")
cm_df$Predicted <- ifelse(cm_df$Predicted == 0, "Negative", "Positive")

# Plot confusion matrix using ggplot2
ggplot(cm_df, aes(x = Actual, y = Predicted, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), vjust = 1) +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  labs(x = "Predicted", y = "Actual", title = "Confusion Matrix") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r}
confusionMatrix(Data$IHC.P53.Status, reference=Data$TP53.Sanger.Status, positive='Positive')
```

## Confusion Matrix at 40% Threshold
```{r}
# Define the threshold
threshold <- 40

# Classify observations based on the threshold
predicted_TP53_status <- ifelse(Data$Percent.Nuclei.Stained.in.p53.IHC >= threshold, 1, 0)

# Create confusion matrix
conf_matrix <- table(Actual = Data$TP53.Sanger.Status..0.1., Predicted = predicted_TP53_status)


conf_matrix_obj <- as.matrix(conf_matrix)

# Create confusion matrix using caret
cm <- confusionMatrix(conf_matrix_obj)

# Convert to data frame for ggplot2
cm_df <- as.data.frame(cm$table)

cm_df$Actual <- ifelse(cm_df$Actual == 0, "Negative", "Positive")
cm_df$Predicted <- ifelse(cm_df$Predicted == 0, "Negative", "Positive")

# Plot confusion matrix using ggplot2
ggplot(cm_df, aes(x = Actual, y = Predicted, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), vjust = 1) +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  labs(x = "Predicted", y = "Actual", title = "Confusion Matrix") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```


## Optimal Cutpoint
```{r}
library(cutpointr)
optimal = cutpointr(Data, Percent.Nuclei.Stained.in.p53.IHC, TP53.Sanger.Status..0.1., method = maximize_metric, metric = sum_sens_spec)
summary(optimal)
```

## ROC Plot

```{r}
# Create ROC curve

basic.roc = ggplot(Data, aes(d=TP53.Sanger.Status, m=Percent.Nuclei.Stained.in.p53.IHC))+geom_roc()
basic.roc
```


## Confidence Intervals

```{r}
# Calculate inverse confidence intervals
conf_intervals <- confint(logmodel)

# Print confidence intervals
print("Confidence intervals for coefficients:")
print(conf_intervals)

# Extract coefficient estimates and their confidence intervals
coef_estimates <- coef(logmodel)
lower_ci <- conf_intervals[, 1]
upper_ci <- conf_intervals[, 2]

# Calculate inverse confidence intervals
inv_lower_ci <- 1 / (1 + exp(-(coef_estimates + lower_ci)))
inv_upper_ci <- 1 / (1 + exp(-(coef_estimates + upper_ci)))

# Create a data frame with inverse confidence intervals
inv_ci_df <- data.frame(coef = names(coef_estimates),
                        inv_lower_ci = inv_lower_ci,
                        inv_upper_ci = inv_upper_ci)

# Print the data frame
print("Inverse Confidence Intervals for Predicted Probabilities:")
print(inv_ci_df)



# Plot inverse confidence intervals
ggplot(inv_ci_df, aes(x = coef, y = inv_lower_ci)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  geom_errorbar(aes(ymin = inv_lower_ci, ymax = inv_upper_ci), width = 0.4) +
  labs(x = "Percent Nuclei Stained in p53 IHC",
       y = "Inverse Confidence Interval",
       title = "Inverse Confidence Intervals on IHC Staining Percentage") +
  theme_minimal()

# Plot bar graph with confidence intervals and point estimates
ggplot(inv_ci_df, aes(x = coef, y = inv_lower_ci)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  geom_errorbar(aes(ymin = inv_lower_ci, ymax = inv_upper_ci), width = 0.4) +
  geom_text(aes(label = round(inv_lower_ci, 2)), vjust = -0.5) +
  labs(x = "Percent Nuclei Stained in p53 IHC",
       y = "Inverse Confidence Interval",
       title = "Inverse Confidence Intervals on IHC Staining Percentage") +
  theme_minimal()



ggplot(inv_ci_df, aes(x=coef, y=frequency(coef)))+geom_point()+
  geom_errorbar(aes(ymin=inv_lower_ci, ymax=inv_upper_ci))


```